Learning and Recognition of \strips\ Action Models from State Observations

### 2018-01-19 doc/ijcai-18/observations-learning.tex
updates to the IJCAI18 paper after evaluation as edition distance proposal

\begin{abstract}
This paper presents a classical planning compilation for learning \strips\ action models from state observations. The compilation approach does not require observing the precise actions that produced the state observations because such actions are determined by an off-the-shelf classical planner. In addition, the compilation is extensible to estimate the probability distribution of the possible \strips\ models given a sequence of state observations. This extension allows us to address model recognition tasks as well as to semantically evaluate the quality of learned \strips\ models.
\end{abstract}

@@

The Machine Learning of planning action models is a promising alternative to hand-coding them and nowadays, there exist sophisticated algorithms like {\sc AMAN}~\cite{zhuo2013action}, {\sc ARMS}~\cite{yang2007learning}, {\sc LOCM}~\cite{cresswell2013acquiring} or {\sc SLAF}~\cite{amir:alearning:JAIR08}. Motivated by recent advances on the synthesis of different kinds of generative models with classical planning~\cite{bonet2009automatic,segovia2016hierarchical,segovia2017generating}, this paper presents a novel approach for learning \strips\ action models that introduces the following contributions:
\begin{enumerate}
\item Is defined as a planning compilation. This fact opens the door to the {\em bootstrapping} of planning action models and allows us to report results over a wide range of planning domains. 
\item Does not require observing the particular executed actions. An off-the-shelf classical planner determines these actions given the state observations.
\item Can estimate the probability distribution of the possible \strips\ models, given an observation sequence. This extension allow us to address model recognition tasks as well as to evaluate the quality of the learned models semantically, without comparing with a reference model.
\end{enumerate}

\section{Background}
This section defines the planning models used in this work as well as the input (states and state invariants) and the output (an \strips\ action model) of the addressed learning task.

@@

An action $a\in A$ is now defined as a set of {\em preconditions} $\pre(a)\in\mathcal{L}(F)$ and a set of {\em conditional effects} $\cond(a)$. Each conditional effect $C\rhd E\in\cond(a)$ is composed of two sets of literals $C\in\mathcal{L}(F)$, the {\em condition}, and $E\in\mathcal{L}(F)$, the {\em effect}. An action $a\in A$ is {\em applicable} in a state $s$ if and only if $\pre(a)\subseteq s$, and the {\em triggered effects} resulting from the action application are the effects whose conditions hold in $s$:
\[
triggered(s,a)=\bigcup_{C\rhd E\in\cond(a),C\subseteq s} E,
\]

@@

Finally we define $F_v(\xi)\subseteq F_v$ as the subset of elements that can appear in a given action schema $\xi$. For instance, for the {\em stack} action schema $F_v({\tt stack})=F_v$ while $F_v({\tt pickup})$={\small\tt\{handempty, holding($v_1$), clear($v_1$), ontable($v_1$), on($v_1,v_1$)\}} only contains the fluents from $F_v$ that do not involve $v_2$.

\begin{figure}
\begin{scriptsize}
\begin{verbatim}
(:action stack
 :parameters (?v1 ?v2 - object)
 :precondition (and (holding ?v1) (clear ?v2))
  :effect (and (not (holding ?v1)) (not (clear ?v2))
               (handempty) (clear ?v1) (on ?v1 ?v2)))
\end{verbatim}
\end{scriptsize}
 \caption{\small \strips\ operator schema coding, in PDDL, the {\em stack} action from the {\em blocksworld}.}
\label{fig:stack}
\end{figure}

@@

We formalize a more challenging learning task, where less input knowledge is available. The addressed learning task corresponds to observing an agent acting in the world but watching only the results of its plan executions, the actual executed actions are unobserved. This learning task is defined as $\Lambda=\tup{\Psi,\Xi,\mathcal{O}}$:
\begin{itemize}
\item $\Psi$ is the set of predicates that define the abstract state space of a given classical planning frame.
\item $\Xi$ is the set of empty operator schemes containing only the headers with the name and parameter of each operator schema $\xi\in\Xi$.
\item $\mathcal{O}=\{s_0,s_1,\ldots,s_{n}\}$ is a sequence of {\em state observations} obtained observing the execution of an {\em unobserved} plan $\pi=\tup{a_1, \ldots, a_n}$.
\end{itemize}

A solution to $\Lambda$ is a set of operator schema $\Xi'$ compliant with the predicates in $\Psi$, the headers in $\Xi$ and the sequence of state observations $\mathcal{O}$. A planning compilation is a suitable approach for addressing a $\Lambda$ learning task because a solution must not only determine the \strips\ action model $\Xi'$ but also, the {\em unobserved} plan $\pi=\tup{a_1, \ldots, a_n}$, that can explain $\mathcal{O}$. Figure~\ref{fig:lexample} shows an example of a $\Lambda$ task for learning a \strips\ action model in the blocksworld from the sequence of five state observations that corresponds to inverting a 2-blocks tower.

\begin{figure}
{\footnotesize\tt ;;; Predicates in $\Psi$}
\begin{scriptsize}
\begin{verbatim}
(handempty) (holding ?o  - object)
(clear ?o - object) (ontable ?o - object)
(on ?o1 - object ?o2 - object)
\end{verbatim}
\end{scriptsize}
\vspace{0.2cm}
{\footnotesize\tt ;;; Headers in $\Xi$}
\begin{scriptsize}
\begin{verbatim}
(pickup v1) (putdown v1) (stack v1 v2} (unstack v1 v2)
\end{verbatim}
\end{scriptsize}
\vspace{0.2cm}
{\footnotesize\tt ;;; Observations in $\mathcal{O}$}
\begin{scriptsize}
\begin{verbatim}
;;; observation #0
(clear block2) (on block2 block1) (ontable block1) 
(handempty)

@@

;;; observation #4
(clear block1) (on block1 block2) (ontable block2) 
(handempty) 
\end{verbatim}
\end{scriptsize}
 \caption{\small Example of a task for learning a \strips\ action model in the blocksworld from a sequence of five state observations.}
\label{fig:lexample}
\end{figure}

@@

Our approach for addressing the learning task, is compiling it into a classical planning task with conditional effects. The intuition behind the compilation is that a solution to the resulting classical planning task is a sequence of actions that:
\begin{enumerate}
\item Programs the \strips\ action model $\Xi'$. A solution plan starts with a {\em prefix} that, for each $\xi\in\Xi$, determines which fluents $f\in F_v$ belong to its $pre(\xi)$, $del(\xi)$ and $add(\xi)$ sets.
\item Validates the \strips\ action model $\Xi'$ in $\mathcal{O}$. The solution plan continues with a postfix that produces the sequence of states in $\mathcal{O}$ starting from $s_0$, and using the programmed action model $\Xi'$. 
\end{enumerate}

Given a learning task $\Lambda=\tup{\Psi,\Xi,\mathcal{O}}$ the compilation outputs a classical planning task $P_{\Lambda}=\tup{F_{\Lambda},A_{\Lambda},I_{\Lambda},G_{\Lambda}}$:
\begin{itemize}
\item $F_{\Lambda}$ contains:
\begin{itemize}
\item Fluents $F$ built instantiating the predicates $\Psi$ with the objects appearing in the input observations $\mathcal{O}$. 
\item Fluents representing the programmed action model $pre_f(\xi)$, $del_f(\xi)$ and $add_f(\xi)$, for every $f\in F_v(\xi)$. If a fluent $pre_f(\xi)/del_f(\xi)/add_f(\xi)$ holds, it means that $f$ is a precondition/negative effect/positive effect in the \strips\ operator schema $\xi\in \Xi$. For instance, the preconditions of the $stack$ schema (Figure~\ref{fig:stack}) are represented by fluents {\small\tt pre\_holding\_stack\_$v_1$} and {\small\tt pre\_clear\_stack\_$v_2$} set to {\em True}.
\item Fluent $mode_{prog}$ indicating whether the operator schemes are programmed or validated (already programmed) and fluents $\{test_i\}_{1\leq i\leq n}$, indicating the observation where the action model is validated.
\end{itemize}
\item $I_{\Lambda}$ contains the fluents from $F$ that encode $s_0$ (the first observation) and every $pre_f(\xi)\in F_{\Lambda}$ and $mode_{prog}$ set to true. Our compilation assumes that initially operator schemes are programmed with every possible precondition, no negative effect and no positive effect.
\item $G_{\Lambda}=\bigcup_{1\leq i\leq n}\{test_i\}$, indicates that the programmed action model is validated in all the input observations.
\item $A_{\Lambda}$ comprises three kinds of actions:
\begin{enumerate}
\item Actions for {\em programming} operator schema $\xi\in\Xi$:
\begin{itemize}
\item Actions for {\bf removing} a {\em precondition} $f\in F_v(\xi)$ from the action schema $\xi\in\Xi$.

@@

\item Actions for {\em applying} an already programmed operator schema $\xi\in\Xi$ bound with the objects $\omega\subseteq\Omega^{ar(\xi)}$. We assume operators headers are known so the binding of the operator schema is done implicitly by order of appearance of the action parameters, i.e. variables $pars(\xi)$ are bound to the objects in $\omega$ appearing at the same position. Figure~\ref{fig:compilation} shows the PDDL encoding of the action for applying a programmed operator $stack$.
\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{apply_{\xi,\omega}})=&\{pre_{f}(\xi)\implies p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))},\\
\cond(\mathsf{apply_{\xi,\omega}})=&\{del_{f}(\xi)\}\rhd\{\neg p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))},\\
&\{add_{f}(\xi)\}\rhd\{p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))},\\
&\{mode_{prog}\}\rhd\{\neg mode_{prog}\}.
\end{align*}
\end{small}

\item Actions for {\em validating} an observation {\tt\small $1\leq i\leq n$}. 
\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{validate_{i}})=&s_i\cup\{test_j\}_{j\in 1\leq j<i}\\
&\cup\{\neg test_j\}_{j\in t\leq j\leq n}\cup \{\neg mode_{prog}\},\\
\cond(\mathsf{validate_{i}})=&\{\emptyset\}\rhd\{test_i\}.
\end{align*}
\end{small}
\end{enumerate}
\end{itemize}

@@

\begin{lemma}
Soundness. Any classical plan $\pi$ that solves $P_{\Lambda}$ induces an action model $\Xi$ that solves the learning task $\Lambda$.
\end{lemma}

\begin{proof}[Proof sketch]
\begin{small}
The compilation forces that once the preconditions of an operator schema $\xi \in \Xi$ are programmed, they cannot be altered. The same happens with the positive and negative effects (furthermore, effects can only be programmed after preconditions are programmed). Once operator schemes are programmed they can only be applied because of the $mode_{prog}$ fluent. To solve $P_{\Lambda}$, goals $\{test_i\}$, {\small $1\leq i\leq n$} can only be achieved: executing an applicable sequence of programmed operator schemes that reaches the state $s_i$, starting from $s_0$. If this is achieved for every {\small $1\leq i\leq n$}, it means that the programmed action model $\Xi$ is compliant with the provided input knowledge and hence, solves $\Lambda$.
\end{small}
\end{proof}

\begin{lemma}
Completeness. Any \strips\ action model $\Xi$ computable from $\Lambda=\tup{\Psi,\Xi,\mathcal{O}}$ can be obtained by solving the corresponding classical planning task $P_{\Lambda}$.
\end{lemma}

\begin{proof}[Proof sketch]
\begin{small}
By definition, $F_v(\xi)\subseteq F_\Lambda$ fully captures the set of elements that can appear in a \strips\ action schema $\xi\in\Xi$. Any possible \strips\ action schema built with the fluents in $F_v$ can be computed because the compilation only constrains the application of $\mathsf{apply_{\xi,\omega}}$ actions iff they are not compliant with the $\mathcal{O}$ sequence.
\end{small}
\end{proof}

\begin{figure}
\begin{scriptsize}
\begin{verbatim}
(:action apply_stack
  :parameters (?o1 - object ?o2 - object)
  :precondition
   (and (or (not (pre_on_stack_v1_v1)) (on ?o1 ?o1))
        (or (not (pre_on_stack_v1_v2)) (on ?o1 ?o2))
        (or (not (pre_on_stack_v2_v1)) (on ?o2 ?o1))
        (or (not (pre_on_stack_v2_v2)) (on ?o2 ?o2))
        (or (not (pre_ontable_stack_v1)) (ontable ?o1))
        (or (not (pre_ontable_stack_v2)) (ontable ?o2))
        (or (not (pre_clear_stack_v1)) (clear ?o1))
        (or (not (pre_clear_stack_v2)) (clear ?o2))
        (or (not (pre_holding_stack_v1)) (holding ?o1))
        (or (not (pre_holding_stack_v2)) (holding ?o2))
        (or (not (pre_handempty_stack)) (handempty)))
  :effect
   (and (when (del_on_stack_v1_v1) (not (on ?o1 ?o1)))
        (when (del_on_stack_v1_v2) (not (on ?o1 ?o2)))
        (when (del_on_stack_v2_v1) (not (on ?o2 ?o1)))
        (when (del_on_stack_v2_v2) (not (on ?o2 ?o2)))
        (when (del_ontable_stack_v1) (not (ontable ?o1)))
        (when (del_ontable_stack_v2) (not (ontable ?o2)))
        (when (del_clear_stack_v1) (not (clear ?o1)))
        (when (del_clear_stack_v2) (not (clear ?o2)))
        (when (del_holding_stack_v1) (not (holding ?o1)))
        (when (del_holding_stack_v2) (not (holding ?o2)))
        (when (del_handempty_stack) (not (handempty)))
        (when (add_on_stack_v1_v1) (on ?o1 ?o1))
        (when (add_on_stack_v1_v2) (on ?o1 ?o2))
        (when (add_on_stack_v2_v1) (on ?o2 ?o1))
        (when (add_on_stack_v2_v2) (on ?o2 ?o2))
        (when (add_ontable_stack_v1) (ontable ?o1))
        (when (add_ontable_stack_v2) (ontable ?o2))
        (when (add_clear_stack_v1) (clear ?o1))
        (when (add_clear_stack_v2) (clear ?o2))
        (when (add_holding_stack_v1) (holding ?o1))
        (when (add_holding_stack_v2) (holding ?o2))
        (when (add_handempty_stack) (handempty))
        (when (modeProg) (not (modeProg)))))
\end{verbatim}
\end{scriptsize}
 \caption{\small Action for applying an already programmed schema $stack$ as encoded in PDDL (implications coded as disjunctions).}
\label{fig:compilation}
\end{figure}

\section{Recognizing \strips\ action models}
Inspired by the {\em Plan recognition as planning} approach~\cite{ramirez2009plan} we show that the previous compilation can be extended to estimate the probability distribution of the possible \strips\ models given a sequence of state observations. This extension allows us to address model recognition tasks as well as to semantically evaluate the quality of learned \strips\ models

\subsection{The \strips\ edit distance}
We assume that how well a given \strips\ action model $\Xi$ explains a given sequence of observations $\mathcal{O}$, depends on the amount of edits that one has to introduce to $\Xi$ to produce the sequence of obsevations $\mathcal{O}$. If the given model $\Xi$ perfectly explains the observations $\mathcal{O}$, no edit have to be introduced and the probability of explanning the observations is $P(\Xi|\mathcal{O})=1$. On the other hand, with a probability of explanning the observations of $P(\Xi|\mathcal{O})=0$, is the model that requires the maximum number of edits to explain the sequence of observations $\mathcal{O}$.

To compute the $P(\Xi|\mathcal{O})$ probability distribution, we need a metric that quantifies this amount of edits. To define such metric, we need to define first which are the possible edit operations that can be performed on a \strips\ action model. With the aim of keeping tractable the branching factor of the classical planning task resulting from our compilation, we only define two \strips\ edit operations:
\begin{itemize}
\item {\em Deletion}. A fluent $pre_f(\xi)/del_f(\xi)/add_f(\xi)$ is removed from the operator schema $\xi$
\item {\em Insertion}. A fluent $pre_f(\xi)/del_f(\xi)/add_f(\xi)$ is added to the operator schema $\xi$
\end{itemize}

Now we are ready to define a metric that asseses then how dissimilar two given \strips\ action models are
\begin{definition}
The {\bf \strips\ edit distance}, denoted as $\delta(\Xi,\Xi')$, is the minimum number of edit operations that allow to transform a given \strips\ action model $\Xi$ into the given \strips\ action model $\Xi'$. This edit distance satisfies the {\em metric axioms} provided that the two operations has the same positive cost.
\end{definition}

>>>>

\begin{abstract}
This paper presents a classical planning compilation for learning \strips\ action models from state observations. The compilation approach does not require observing the precise actions that produced the state observations because such actions are determined by a classical planner. In addition, the compilation is extensible to (1) semantically evaluate the quality of learned \strips\ models and (2), estimate the probability distribution of the possible \strips\ models given a sequence of state observations.  
\end{abstract}

@@

The Machine Learning of planning action models is a promising alternative to hand-coding them and nowadays, there exist sophisticated algorithms like {\sc AMAN}~\cite{zhuo2013action}, {\sc ARMS}~\cite{yang2007learning}, {\sc LOCM}~\cite{cresswell2013acquiring} or {\sc SLAF}~\cite{amir:alearning:JAIR08}. Motivated by recent advances on the synthesis of different kinds of generative models with classical planning~\cite{bonet2009automatic,segovia2016hierarchical,segovia2017generating}, this paper presents a novel approach for learning \strips\ action models that introduces the following contributions:
\begin{enumerate}
\item Is defined as a classical planning compilation. This fact opens the door to the {\em bootstrapping} of planning action models and allows us to report results over a wide range of planning domains. 
\item Does not require observing the particular executed actions. An off-the-shelf classical planner determines these actions given the state observations.
\item Is extensible to semantically evaluate the quality of learned \strips\ models as well as to estimate the probability distribution of the possible \strips\ models given a sequence of state observations. 
\end{enumerate}

\section{Background}
This section defines the planning models used in this work as well as the output (an \strips\ action model) of the addressed learning task.

@@

An action $a\in A$ with conditional effects is defined as a set of {\em preconditions} $\pre(a)\in\mathcal{L}(F)$ and a set of {\em conditional effects} $\cond(a)$. Each conditional effect $C\rhd E\in\cond(a)$ is composed of two sets of literals $C\in\mathcal{L}(F)$, the {\em condition}, and $E\in\mathcal{L}(F)$, the {\em effect}. An action $a\in A$ is {\em applicable} in a state $s$ if and only if $\pre(a)\subseteq s$, and the {\em triggered effects} resulting from the action application are the effects whose conditions hold in $s$:
\[
triggered(s,a)=\bigcup_{C\rhd E\in\cond(a),C\subseteq s} E,
\]

@@

Finally we also define $F_v(\xi)\subseteq F_v$ as the subset of elements that can appear in a given action schema $\xi$. For instance, for the {\em stack} action schema $F_v({\tt stack})=F_v$ while $F_v({\tt pickup})$={\small\tt\{handempty, holding($v_1$), clear($v_1$), ontable($v_1$), on($v_1,v_1$)\}} only contains the fluents from $F_v$ that do not involve $v_2$ because the action header contained the single parameter $v_1$.

\begin{figure}
\begin{footnotesize}
\begin{verbatim}
(:action stack
 :parameters (?v1 ?v2 - object)
 :precondition (and (holding ?v1) (clear ?v2))
  :effect (and (not (holding ?v1)) 
               (not (clear ?v2))
               (handempty) (clear ?v1) 
               (on ?v1 ?v2)))
\end{verbatim}
\end{footnotesize}
 \caption{\small \strips\ operator schema coding, in PDDL, the {\em stack} action from the {\em blocksworld}.}
\label{fig:stack}
\end{figure}

@@

In this paper we address a more challenging learning task, where less input knowledge is available. The addressed learning task corresponds to observing an agent acting in the world but watching only the results of its plan executions, the actual executed actions are unobserved. This learning task is defined as $\Lambda=\tup{\Psi,\Xi,\mathcal{O}}$, where:
\begin{itemize}
\item $\Psi$ is the set of predicates that define the abstract state space of a given classical planning frame.
\item $\Xi$ is the set of empty operator schemes containing only the headers with the name and parameter of each operator schema $\xi\in\Xi$.
\item $\mathcal{O}=\tup{s_0,s_1,\ldots,s_{n}}$ is a sequence of {\em state observations} obtained observing the execution of an {\em unobserved} plan $\pi=\tup{a_1, \ldots, a_n}$.
\end{itemize}

A solution to $\Lambda$ is a set of operator schema $\Xi'$ compliant with the predicates in $\Psi$, the headers in $\Xi$ and the given sequence of state observations $\mathcal{O}$. A planning compilation is a suitable approach for addressing a $\Lambda$ learning task because a solution must not only determine the \strips\ action model $\Xi'$ but also, the {\em unobserved} plan $\pi=\tup{a_1, \ldots, a_n}$, that explains $\mathcal{O}$. Figure~\ref{fig:lexample} shows an example of a $\Lambda$ task for learning a \strips\ action model in the blocksworld from the sequence of five state observations that corresponds to inverting a 2-blocks tower.

\begin{figure}[hbt!]
{\footnotesize\tt ;;; Predicates in $\Psi$}
\begin{footnotesize}
\begin{verbatim}
(handempty) (holding ?o  - object)
(clear ?o - object) (ontable ?o - object)
(on ?o1 - object ?o2 - object)
\end{verbatim}
\end{footnotesize}
\vspace{0.2cm}
{\footnotesize\tt ;;; Headers in $\Xi$}
\begin{footnotesize}
\begin{verbatim}
(pickup v1) (putdown v1) 
(stack v1 v2} (unstack v1 v2)
\end{verbatim}
\end{footnotesize}
\vspace{0.2cm}
{\footnotesize\tt ;;; Observations in $\mathcal{O}$}
\begin{footnotesize}
\begin{verbatim}
;;; observation #0
(clear block2) (on block2 block1) 
(ontable block1) (handempty)

@@

;;; observation #4
(clear block1) (on block1 block2) 
(ontable block2) (handempty) 
\end{verbatim}
\end{footnotesize}
 \caption{\small Example of a $\Lambda$ task for learning a \strips\ action model in the {\em blocksworld} from a sequence of five state observations.}
\label{fig:lexample}
\end{figure}

@@

Our approach for addressing the learning task, is compiling it into a classical planning task with conditional effects. The intuition behind the compilation is that a solution to the resulting classical planning task is a sequence of actions that:
\begin{enumerate}
\item Programs the \strips\ action model $\Xi'$. A solution plan starts with a {\em prefix} that, for each $\xi\in\Xi$, determines which fluents $f\in F_v(\xi)$ belong to its $pre(\xi)$, $del(\xi)$ and $add(\xi)$ sets.
\item Validates the \strips\ action model $\Xi'$ in $\mathcal{O}$. The solution plan continues with a postfix that produces the given sequence of states $\tup{s_0,s_1,\ldots,s_{n}}$ using the programmed action model $\Xi'$. 
\end{enumerate}

Given a learning task $\Lambda=\tup{\Psi,\Xi,\mathcal{O}}$ the compilation outputs a classical planning task $P_{\Lambda}=\tup{F_{\Lambda},A_{\Lambda},I_{\Lambda},G_{\Lambda}}$:
\begin{itemize}
\item $F_{\Lambda}$ contains:
\begin{itemize}
\item The set of fluents $F$ that is built instantiating the predicates $\Psi$ with the objects appearing in the input observations $\mathcal{O}$. 
\item Fluents representing the programmed action model $pre_f(\xi)$, $del_f(\xi)$ and $add_f(\xi)$, for every $f\in F_v(\xi)$. If a fluent $pre_f(\xi)/del_f(\xi)/add_f(\xi)$ holds, it means that $f$ is a precondition/negative effect/positive effect in the \strips\ operator schema $\xi\in \Xi$. For instance, the preconditions of the $stack$ schema (Figure~\ref{fig:stack}) are represented by the fluents {\small\tt pre\_holding\_stack\_$v_1$} and {\small\tt pre\_clear\_stack\_$v_2$} set to {\em True}.
\item Fluent $mode_{prog}$ indicating whether the operator schemes are programmed or validated (already programmed) and fluents $\{test_i\}_{1\leq i\leq n}$, indicating the observation where the action model is validated.
\end{itemize}
\item $I_{\Lambda}$ contains the fluents from $F$ that encode $s_0$ (the first observation) and $mode_{prog}$ set to true. In addition, our compilation assumes that initially operator schemes are programmed with every possible precondition, no negative effect and no positive effect. With this regard, the fluents $pre_f(\xi)$ hold at the initial state for every $f\in F_v(\xi)$.
\item $G_{\Lambda}=\bigcup_{1\leq i\leq n}\{test_i\}$, indicates that the programmed action model is validated in all the input observations.
\item $A_{\Lambda}$ comprises three kinds of actions:
\begin{enumerate}
\item Actions for {\em programming} operator schema $\xi\in\Xi$:
\begin{itemize}
\item Actions for {\bf removing} a {\em precondition} $f\in F_v(\xi)$ from the action schema $\xi\in\Xi$.

@@

\item Actions for {\em applying} an already programmed operator schema $\xi\in\Xi$ bound with the objects $\omega\subseteq\Omega^{ar(\xi)}$. We assume that the operators headers are known so the binding of the operator schema is done implicitly by order of appearance of the action parameters, i.e. variables $pars(\xi)$ are bound to the objects in $\omega$ appearing at the same position. Figure~\ref{fig:compilation} shows the PDDL encoding of the action for applying a programmed operator $stack$.
\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{apply_{\xi,\omega}})=&\{pre_{f}(\xi)\implies p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))},\\
\cond(\mathsf{apply_{\xi,\omega}})=&\{del_{f}(\xi)\}\rhd\{\neg p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))},\\
&\{add_{f}(\xi)\}\rhd\{p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))},\\
&\{mode_{prog}\}\rhd\{\neg mode_{prog}\}.
\end{align*}
\end{small}

\item Actions for {\em validating} an observation {\tt\small $1\leq i\leq n$}. 
\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{validate_{i}})=&s_i\cup\{test_j\}_{j\in 1\leq j<i}\\
&\cup\{\neg test_j\}_{j\in i\leq j\leq n}\cup \{\neg mode_{prog}\},\\
\cond(\mathsf{validate_{i}})=&\{\emptyset\}\rhd\{test_i\}.
\end{align*}
\end{small}
\end{enumerate}
\end{itemize}

\begin{figure}[hbt!]
\begin{scriptsize}
\begin{verbatim}
(:action apply_stack
  :parameters (?o1 - object ?o2 - object)
  :precondition
   (and (or (not (pre_on_stack_v1_v1)) (on ?o1 ?o1))
        (or (not (pre_on_stack_v1_v2)) (on ?o1 ?o2))
        (or (not (pre_on_stack_v2_v1)) (on ?o2 ?o1))
        (or (not (pre_on_stack_v2_v2)) (on ?o2 ?o2))
        (or (not (pre_ontable_stack_v1)) (ontable ?o1))
        (or (not (pre_ontable_stack_v2)) (ontable ?o2))
        (or (not (pre_clear_stack_v1)) (clear ?o1))
        (or (not (pre_clear_stack_v2)) (clear ?o2))
        (or (not (pre_holding_stack_v1)) (holding ?o1))
        (or (not (pre_holding_stack_v2)) (holding ?o2))
        (or (not (pre_handempty_stack)) (handempty)))
  :effect
   (and (when (del_on_stack_v1_v1) (not (on ?o1 ?o1)))
        (when (del_on_stack_v1_v2) (not (on ?o1 ?o2)))
        (when (del_on_stack_v2_v1) (not (on ?o2 ?o1)))
        (when (del_on_stack_v2_v2) (not (on ?o2 ?o2)))
        (when (del_ontable_stack_v1) (not (ontable ?o1)))
        (when (del_ontable_stack_v2) (not (ontable ?o2)))
        (when (del_clear_stack_v1) (not (clear ?o1)))
        (when (del_clear_stack_v2) (not (clear ?o2)))
        (when (del_holding_stack_v1) (not (holding ?o1)))
        (when (del_holding_stack_v2) (not (holding ?o2)))
        (when (del_handempty_stack) (not (handempty)))
        (when (add_on_stack_v1_v1) (on ?o1 ?o1))
        (when (add_on_stack_v1_v2) (on ?o1 ?o2))
        (when (add_on_stack_v2_v1) (on ?o2 ?o1))
        (when (add_on_stack_v2_v2) (on ?o2 ?o2))
        (when (add_ontable_stack_v1) (ontable ?o1))
        (when (add_ontable_stack_v2) (ontable ?o2))
        (when (add_clear_stack_v1) (clear ?o1))
        (when (add_clear_stack_v2) (clear ?o2))
        (when (add_holding_stack_v1) (holding ?o1))
        (when (add_holding_stack_v2) (holding ?o2))
        (when (add_handempty_stack) (handempty))
        (when (modeProg) (not (modeProg)))))
\end{verbatim}
\end{scriptsize}
 \caption{\small Action for applying an already programmed schema $stack$ as encoded in PDDL (implications coded as disjunctions).}
\label{fig:compilation}
\end{figure}

@@

\begin{lemma}
Soundness. Any classical plan $\pi$ that solves $P_{\Lambda}$ induces an action model $\Xi'$ that solves the learning task $\Lambda$.
\end{lemma}

\begin{proof}[Proof sketch]
\begin{small}
The compilation forces that once the preconditions of an operator schema $\xi \in \Xi'$ are programmed, they cannot be altered. The same happens with the positive and negative effects. Furthermore because of the preconditions of the $\mathsf{programPre_{f,\xi}}$ actions, effects are only programmable after preconditions are programmed. Once the operator schemes $\Xi'$ are programmed, they can only be applied because of the $mode_{prog}$ fluent. To solve $P_{\Lambda}$, the goals $\{test_i\}$, {\small $1\leq i\leq n$} can only be achieved: executing an applicable sequence of programmed operator schemes that reaches every state $s_i\in\mathcal{O}$, starting from $s_0$ and following  the sequence {\small $1\leq i\leq n$}. Therefore if $test_n$ is achieved, it means that the programmed action model $\Xi'$ is compliant with the provided input knowledge and hence, solves $\Lambda$.
\end{small}
\end{proof}

\begin{lemma}
Completeness. Any \strips\ action model $\Xi'$ that solves a $\Lambda=\tup{\Psi,\Xi,\mathcal{O}}$ learning task, is computable solving the corresponding classical planning task $P_{\Lambda}$.
\end{lemma}

\begin{proof}[Proof sketch]
\begin{small}
By definition, $F_v(\xi)\subseteq F_\Lambda$ fully captures the set of elements that can appear in a \strips\ action schema $\xi\in\Xi$. In addition, any possible \strips\ action schema $\Xi'$ that can be built with the fluents in $F_v$ can be computed  with the $P_{\Lambda}$ compilation. The only \strips\ action models that the compilation cannot compute are the ones that violate the state constraints defined by the $\mathcal{O}$ sequence, which constrain as well the solutions to the $\Lambda$ learning task.
\end{small}
\end{proof}

\section{Recognizing \strips\ action models}
Inspired by {\em plan recognition as planning}~\cite{ramirez2009plan}, we show that the previous compilation can be extended to (1) semantically evaluate the quality of learned \strips\ models, without comparing with a reference model and (2), address \strips\ action model recognition tasks.

\subsection{The \strips\ edit distance}
We assume that how well a given \strips\ action model $\Xi$ explains a given sequence of observations $\mathcal{O}$, depends on the amount of {\em edition} that one has to introduce to $\Xi$ to produce the sequence of obsevations $\mathcal{O}$. In the extreme, if the given model $\Xi$ perfectly explains the observations $\mathcal{O}$, no {\em edit} to the model must be introduced.

Let us define the operations allowed to edit a given \strips\ action model. We call these operations the {\em \strips\ edit operations}. With the aim of keeping tractable the branching factor of the classical planning task that results from our compilation, we only define two {\em \strips\ edit operations}:
\begin{itemize}
\item {\em Deletion}. A fluent $pre_f(\xi)/del_f(\xi)/add_f(\xi)$, such that $f\in F_v(\xi)$, is removed from the operator schema $\xi\in\Xi$.
\item {\em Insertion}. A fluent $pre_f(\xi)/del_f(\xi)/add_f(\xi)$, such that $f\in F_v(\xi)$, is added to the operator schema $\xi\in\Xi$.
\end{itemize}

With this defined, we can now formalize an edit distance metric that assesses how dissimilar two given \strips\ action models are. This distance satisfies the {\em metric axioms} provided that the two {\em \strips\ edit operations} have the same positive cost.
\begin{definition}
Given two \strips\ action model $\Xi$ and $\Xi'$, such that both are built from the same set of possible elements $F_v$. The {\em \bf \strips\ edit distance}, denoted as $\delta(\Xi,\Xi')$, is the minimum number of {\em \strips\ edit operations} to transform $\Xi$ into $\Xi'$.
\end{definition}

Since the size of the $F_v$ set is bound, the maximum number of edits that can be introduced to a given \strips\ action model is bound as well. In more detail, for a given operator schema $\xi\in\Xi$ the maximum number of edits that can be introduced to their precondition set is $|F_v(\xi)|$. With regard to effects of $\xi$, the maximum number of edits that can be introduced are two times $|F_v(\xi)|$.

\begin{definition}
Given a \strips\ action model $\Xi$ built from the set of possible elements $F_v$. The {\em \bf maximum  \strips\ edit distance} is the upperbound $\delta(\Xi,*)=\sum_{\xi\in\Xi} 3|F_v(\xi)|$, that indicates the maximum value of the {\em \strips\ edit distance} from $\Xi$ to any \strips\ action model definable within $F_v$. 
\end{definition}

Likewise the {\em \strips\ edit distance} can be defined with regard to a sequence of observations $\mathcal{O}$.
\begin{definition}
 Given a \strips\ action model $\Xi$ built from the set of possible elements $F_v$ and a sequence of observations $\mathcal{O}$. The {\em \bf observations  \strips\ edit distance}, denoted by  $\delta(\Xi,\mathcal{O})$, is the minimum number of {\em \strips\ edit operations} to transform $\Xi$ into an action model $\Xi'$ defined within $F_v$ that can produce a plan $\pi=\tup{a_1, \ldots, a_n}$ such that $\pi$ induces the state observations $\mathcal{O}=\tup{s_0, s_1, \ldots, s_n}$.
\end{definition}

The {\em observations  \strips\ edit distance} allow us to semantically evaluate the quality of a learned strips action model with respect to a sequence of observations that acts as a test set. The lower this distance is, the better the quality of the learned model. The semantic nature of this evaluation is robust to learning episodes where actions are reformulated and still compliant with the inputs (e.g. the {\em blocksworld} operator {\small\tt stack} could be {\em learned} with the preconditions and effects of the {\small\tt unstack} operator and vice versa or the roles of actions parameters with the same type could be interchanged).

\subsection{Recognition of \strips\ models}
Given a set of possible \strips\ models and a sequence of state observations the {\em recognition of \strips\ models} is the task of computing the probability distribution of these models according to the given observations. Here we show that the {\em observations  \strips\ edit distance} is helpful to compute the probability distribution of the possible \strips\ models given a sequence of state observations. The main idea, taken from {\em plan recognition as planning}~\cite{ramirez2009plan}, is to map these distances into likelihoods using the Bayes rule.
